Simplifying Lotka--Volterra interaction matrices by species grouping: optimal groupings are unpredictable

### 2022-05-22 article.rnw
Remove a commented out paragraph break.
Correct spelling.

It is commonly assumed that species which are similar in their responses to competition would automatically be similar in their competition effects.
However, competitive responses may be governed by a different set of traits than compeititive effects, so this assumption does not necessarily hold.
We avoided this assumption by allowing species which were similar in one regard to be dissimilar in the other, and vice versa.
As such, each model was defined by two groupings of species: one grouping of the competitive responses, and one of the competitive effects.

To better understand the meaning of species grouping of competition coefficients, it is useful to recognize that this component of the Lotka--Volterra model can be represented as an interaction matrix \parencite{novak2016characterizing}, where a row contains all the competitive effects on a single species and a column contains all the competitive effects of a single species.
As such, if two species are assumed to have the same response to competition, then two rows of the interaction matrix are constrained to be equal.
Similarly, if two species are assumed to have the same competitive effect, then two columns of the interaction matrix are constrained to be equal.
This allows us to refer below to the grouping of the competitive responses as the ``row grouping'' and the grouping of the competitive effects as the ``column grouping''.
Formally, if species $a$ and $b$ were in the same row group, $\alpha_{aj}\equiv\alpha_{bj} \, \forall j$, and if $a$ and $b$ were in the same column group, $\alpha_{ia}\equiv\alpha_{ib} \, \forall i$.

@@

However, we could test the row-grouping and column-grouping coclassification matrices against one another.
By testing for correlation between the coclassification matrix for the the rows and that for the columns, we could perform a different test regarding the assumption that it is appropriate to use the same grouping for both.
We did this using a Mantel test, as appropriate for similarity matrices.
 %
We could also test the coclassification matrices for correlation with an expected dissimilarity matrix.
We calculated one such dissimilarity matrix based on a species trait: body size (as mean wet mass of female adults, obtained from \cite{terry2021natural}).
The trait difference matrix was calculated as absolute difference between trait values, while the coclassification matrices were converted to dissimilarity matrices by subtracting them from $1$.
We then performed Mantel tests for correlation between the trait distance matrix and each of the two coclassification dissimilarity matrices.

@@

One possible explanation is that species simply do not group to the same extent in a competitive context.
That said, the results suggest that this is unlikely as grouping was still likely to increase parsimony.
Further, there is no reason to suspect that species would not have clear similarities or dissimilarities in their ecological niches, and therefore their competitive effects and responses \parencite{XXX}.
In the present case, we cannot rule out the possibility that a lack of clarity in the groupings is instead a product of the study system.
All the species in this study were members of the same genus, \emph{Drosophila}, and as such were all rather similar.
A study system with species spanning multiple functional groups, or a broader taxonomy, might have shown clearer groupings.
Alternatively, the small number of species in the study might not have been sufficient for clear groups to be defined.
Alas, the brute force method we employed in this study has a very poor algorithmic complexity, preventing the application of the method to studies with more than seven or eight species. %DBS: Add in the number of models for both of these to emphasize the point.
As such, a better method would be required in order to test for grouping of species interactions among a larger set of species.

>>>>

It is commonly assumed that species which are similar in their responses to competition would automatically be similar in their competition effects.
However, competitive responses may be governed by a different set of traits than competitive effects, so this assumption does not necessarily hold.
We avoided this assumption by allowing species which were similar in one regard to be dissimilar in the other, and vice versa.
As such, each model was defined by two groupings of species: one grouping of the competitive responses, and one of the competitive effects.

To better understand the meaning of species grouping of competition coefficients, it is useful to recognise that this component of the Lotka--Volterra model can be represented as an interaction matrix \parencite{novak2016characterizing}, where a row contains all the competitive effects on a single species and a column contains all the competitive effects of a single species.
As such, if two species are assumed to have the same response to competition, then two rows of the interaction matrix are constrained to be equal.
Similarly, if two species are assumed to have the same competitive effect, then two columns of the interaction matrix are constrained to be equal.
This allows us to refer below to the grouping of the competitive responses as the ``row grouping'' and the grouping of the competitive effects as the ``column grouping''.
Formally, if species $a$ and $b$ were in the same row group, $\alpha_{aj}\equiv\alpha_{bj} \, \forall j$, and if $a$ and $b$ were in the same column group, $\alpha_{ia}\equiv\alpha_{ib} \, \forall i$.

@@

However, we could test the row-grouping and column-grouping coclassification matrices against one another.
By testing for correlation between the coclassification matrix for the the rows and that for the columns, we could perform a different test regarding the assumption that it is appropriate to use the same grouping for both.
We did this using a Mantel test, as appropriate for similarity matrices.
We could also test the coclassification matrices for correlation with an expected dissimilarity matrix.
We calculated one such dissimilarity matrix based on a species trait: body size (as mean wet mass of female adults, obtained from \cite{terry2021natural}).
The trait difference matrix was calculated as absolute difference between trait values, while the coclassification matrices were converted to dissimilarity matrices by subtracting them from $1$.
We then performed Mantel tests for correlation between the trait distance matrix and each of the two coclassification dissimilarity matrices.

@@

One possible explanation is that species simply do not group to the same extent in a competitive context.
That said, the results suggest that this is unlikely as grouping was still likely to increase parsimony.
Further, there is no reason to suspect that species would not have clear similarities or dissimilarities in their ecological niches, and therefore their competitive effects and responses \parencite{XXX}.
In the present case, we cannot rule out the possibility that a lack of clarity in the groupings is instead a product of the study system.
All the species in this study were members of the same genus, \emph{Drosophila}, and as such were all rather similar.
A study system with species spanning multiple functional groups, or a broader taxonomy, might have shown clearer groupings.
Alternatively, the small number of species in the study might not have been sufficient for clear groups to be defined.
Alas, the brute force method we employed in this study has a very poor algorithmic complexity, preventing the application of the method to studies with more than seven or eight species. %DBS: Add in the number of models for both of these to emphasise the point.
As such, a better method would be required in order to test for grouping of species interactions among a larger set of species.
